Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-06-08 ICML/ARTEMISS_proposal.tex
Version for feedback Gerko.

To draw inference from incomplete data, most imputation software packages use iterative imputation procedures. With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data?

@@

Our results demonstrate that inferential validity is achieved after five to ten iterations---even in the worst missingness conditions (i.e., up-to 95\% of cases having missing values). Conditions with a lower proportion of incomplete cases yield valid inferences almost instantly (after just two iterations). This is in grave contrast to the non-convergence identified by the diagnostic methods. Autocorrelation- and $\widehat{R}$-values still indicate improving convergence after 20 to 30 iterations.

@@

For example, in Figure \ref{fig:res} we see that... The novel parameter has equal performance to the quantity of scientific interest as parameter ($r=.975$), while having the advantage of being model-independent.

>>>>

To draw inference from incomplete data, most imputation software packages use iterative imputation procedures. With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained using the imputed data?

@@

Our results demonstrate that inferential validity is achieved after five to ten iterations, even under severe missingness conditions (i.e., up-to 95\% of cases having missing values). Conditions with a lower proportion of incomplete cases yield valid inferences almost instantly---after just two iterations. This is in grave contrast to the non-convergence identified by the diagnostic methods. Autocorrelation- and $\widehat{R}$-values still indicate improving convergence after 20 to 30 iterations.

@@

For example, in Figure \ref{fig:res} we see...

The novel parameter has equal performance to the quantity of scientific interest as parameter (Spearman $r=.978$), while having the advantage of being model-independent.
